{\bf \Large More than Meets the Eye: The Power of Context in Decision-Making and Recommendations}

### 2023-06-02 draft.tex
minor changes ex-post
rewrote conclusion

\title{{\bf \Large TO BE RENAMED}
}

@@

    This thesis demonstrates the existence of context effects in multi-attribute, multidimensional settings and develops methodologies on enhancing recommender systems design with context effects. Additionally, this study investigates external context, namely, the tools that allows users to express their preferences and how to make recommender systems better through them.

\section{Introduction}

@@

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates and uses two of them. The first one posits that choice context \footnote{For the remainder of the paper, I will follow the previous literature \citep{truebloodEtAl13} and refer this definition of choice context as context effects.} is ``the availability and the nature of choice alternatives" \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Context also referred to as ``the time and content of the choice, the location or socio-demographic characteristics of the decision-maker..." \citep{adomavicius2011context}.

@@

Research agrees on importance of metrics beyond accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations which are not only in the category ``exactly what I want", but also ``I never thought I would have liked this one" \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase the satisfaction of users \citep{knijnenburgExplainingUserExperience2012}. Albeit, it was observed that the preferences of individuals are not stable and tend to change and system designers have proposed various tools to enable users signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on context arising from the user side, e.g their preferences \citep{adomavicius2011context}. In my last study I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

@@

\section{Application of mathematical decision making model to observational data}\label{chapter:simulationStudy}
\begin{abstract}

    Previous computational decision making models developed to account for context effects have only been
    studied with an experimental data where only one effect was produced at a time. Using data coming from strictly controlled experimental environments
    hinders the understanding of context effects that occur in real-world choice scenarios where items have multiple dimensions and choice sets have dozens of alternatives. In this chapter
    I apply a computational model to an observational data which was not done before. The data comes from an air travel industry and is ideal to study context effects in multi-attribute, multialternative choice environments. I first find optimal parameters for computational model using differential evolution algorithm. Then, I complement a traditional choice model with its outputs and assess the significance of its contribution. This chapter contributes to context effect and decision making literature by providing further insights on behavior of computational decision making models in real-world choice data.

@@

\subsection{Big three context effects}

@@

\section{Choice modeling with context effects:
Generalization to multi-option and multi-attribute settings\footnote{This chapter is based on a joint work with my supervisor Zakaria Babutsidze, William Rand, Nobuyuki Hanaki, Ismael Rafai, Rodrigo Acuna Agost and Thierry Delahaye.}}\label{chapter:jmrPaper}

@@

Context effects have been systematically studied in marketing and psychology \citep{kivetz04, roodrkerkEtAl11, frederickEtAl14, dotsonEtAl18}. However, virtually all such studies have used controlled experiments in neat, compact settings. Namely, the settings where decision-makers are presented with few options and (very) few attributes across which these options differ. In contrast, most actual choices take place in much messier environments. Especially today when much of our search and shopping activity has shifted online. Proliferation of search engines allows each option to easily be compared with many alternatives across many different characteristics. Despite this, we know very little about the prevalence of context effects in these environments. Precise measurements of context effects are not well-defined in multi-option and multi-attribute settings. Defining how to measure these effects is a minimum requirement for proceeding to evaluate the existence of context effects using observational data.

There have been recent attempts in computer science to define some of these effects. Machine learning community has incorporated context effects in discrete choice models applied to observational data \citep{pfannschmidt2019learning, bowerBalzano20}. However, the aim therein is increased prediction accuracy of choice models \citep{tomlinsonBenson21}. As a result, incorporation of context effects takes the form of generalizing choice models to allow for departure from the strict rationality assumptions \footnote{Recent examples of this approach are  Contextual Multinomial Logit by Yusefi Maragheh et al. \citeyearonly{yousefi2020choice} and Linear Context Logit by Tomlinson and Benson \citeyearonly{tomlinsonBenson21} .}. These proposed generalizations of estimated functional forms usually do not distinguish across various different types of context effects. Additionally, these approaches often run into computational difficulties, i.e., the estimation process is NP-hard \citep{yousefi2020choice}.

In this chapter, I propose measures of different context effects in multi-option and multi-dimensional settings. Following Rooderkerk, Van Heerde, and Bijmolt \citeyearonly{roodrkerkEtAl11}, I consider three context effects - attraction, compromise and similarity effects. Attraction effect refers to the increase in attractiveness of a set of option as a result of adding an alternative to the choice set, compromise effect refers to the inclination of consumers to prefer options that represent a compromise across extreme sets of alternatives, while similarity effect refers to the drop in choice likelihood for an alternative once another, similar, alternative has been added to the choice set. Each of the measures corresponding to three aforementioned effects requires a specific approach for making measurements applicable to observational data. Each of these measures is calculated prior to choice estimation, which avoids computational problems. After presenting the generalized measures of the three effects, we perform an empirical analysis of choices based on the new measures using observational data. We use an extensive dataset of airfare choices for this exercise. We identify that attraction and similarity effects do influence choices in air-travel booking data. We also detect a reverse compromise effect that seems to indicate that air-travelers consistently prefer extreme alternatives (i.e., the cheapest, or the shortest flight) to alternatives that constitute a compromise among extreme options.

@@

\newpage
\section{Use of clustering for consideration set modeling in recommender systems\footnote{This chapter is based on a joint work with my supervisor Zakaria Babutsidze, William Rand and Thierry Delahaye. It has been published in the proceedings of 54th Hawaii International Conference on System Sciences}.}\label{chapter:hicssPaper}

@@

In this paper, I incorporate insights from the decision literature into recommender system design using the dataset of European flight choices \citep{lheritier2019airline}. I argue that the choice process occurs in two sequential stages: consumers first identify the small subset of the choices that they would ``consider" and then make a choice from that subset. Combining findings in marketing, management and consumer behavior, and using clustering to quantify the contextual information of the choice set,  I propose a user-side, two-step  ``consider-then-choose" \citep{liu2011efficient, gilbride2004choice} approach to recommender system design to tackle the cold-start problem.

@@

Elimination by aspects is another heuristic approach which has been
proposed in the literature \citep{tversky1972elimination}. The basic setup of this approach is that an individual chooses one attribute, and eliminates options based on this attribute and repeats this procedure for other attributes if necessary until the remaining options do not share common attributes anymore. Then, as a last step, the final option is chosen according to Luce’s choice axiom \citep{luce2012individual} which states that the probability of selecting one option over the others in a choice set is not affected by presence or absence of other options. Most of the results on this topic \citep{batsell1985new,gensch1987two,manrai1989elimination,currim1988disaggregate} indicate that the use of multi-phase heuristic processes can increase the accuracy of the estimation and result in improved interpretability of the models. Elimination by aspects is considered a heuristic method with stochastic rules because of the nature of the comparisons an individual makes, and because the selection process is not based on the relative importance of the features 
\citep{aribarg2018advancing}. Such models are also hard to apply successfully because they require tremendous numbers of parameters to be estimated \citep{batsell1985new}. Although these models can theoretically capture the essence of the two-stage choice process, they are not able to identify the results of separate stages \citep{gilbride2004choice}.

@@

For an empirical study of consideration set formation, one can elicit information on consideration sets in multiple ways \citep{ding2011unstructured,gaskin2007two,yee2007greedoid}. However, for modeling purposes the literature discusses two main ways of consideration set formation: deterministic \citep{coombs1951mathematical} and stochastic \citep{mcfadden1973conditional, urban1984testing}. While stochastic modeling makes all potential sets possible by attaching non-zero choice probability to each of them, deterministic approaches may render some outcomes impossible \citep{aribarg2018advancing}. Because we can not know which answer is the best and the decision-maker, or consumer, is the final arbiter of the ``correct" choice \citep{hauser2014consideration}, the use of either of these two approaches must consider the choice environment, time frame, future value (or loss) associated with the correct and incorrect choice, and so on \citep{punj2009information}. For example, let's consider the flight booking case again and suppose that the consumer lives far from the airport and can reach it only in the afternoon. Consequently, all tickets with departure time before midday would not be considered at all. When forming a consideration set in this case, one must not only consider the characteristics of available options, but also the characteristics of the consumer and the choice environment. While using a purely stochastic approach might yield sets which include some options that consumer would indeed consider, there will also be options which will have zero probability for consideration. In contrast, applying some deterministic rules derived from this particular choice environment, such as the departure time, in the consideration set formation, will exclude those options completely.

When there are not many options to consider, options have few attributes, or final choice utility is not evenly distributed among the attributes, the consideration sets may be modelled via simple deterministic rules, because there is not much cognitive load and the decision rules are relatively simple  \citep{hauser2014consideration, lee2004effect} . When forming consideration sets, it is also important to consider their size. It is very difficult to decide an optimal size based on the choice environment and individual processing capabilities \citep{de2011modelling}.

@@

\subsection{Conclusion and discussion}

@@

In this study, I demonstrated the importance of context effects derived from choice sets in shaping user decisions, as well as the feasibility of incorporating these effects into the design of recommender systems.
Albeit enhancing the system's ability to align its recommendations on the account of user's two-step decision making process is a significant step forward, one must also recognize that context does not arise from the choice set alone. In previous chapters I have used Tversky's \citeyearonly{tversky1972elimination} definition of context being ``the composition  and the nature of the choice set, availability of various options in it". More recent stream of research has also established that choice context is also dependent on user preferences among other things \citep{adomavicius2011context, dey2001understanding}.

@@

\section{User control and recommender systems acceptance}\label{chapter:UserControlAndRS}
\begin{abstract}

@@

The importance of precisely defining the concepts and constructs we use in our study is crucial. Defining the key constructs not only establishes a common understanding, but also provides boundary conditions and ensures the reliability of the research findings \citep{creswell2013research}.

By clearly defining RS we are able to distinguish among related concepts such as recommendation agents (RA) and interactive decision aid tools (IDA) which wile sharing some commonalities, also entail some important differences in their goals, functionalities and user interactions \citep{xiao2007commerce}. By specifying the characteristics of RS in our particular context, we aim to ensure the relevance of our findings. This is also important considering our methodology and framework.

@@

Henceforth, in the following subsections, we will discuss the definitions of recommender systems and user control in detail and justify our choices for this study.

@@

The second stage of user control entails functionalities which allow users to influence recommendations after they are generated. The previous approaches have user sorting, feature and weight based filtering and critiquing. \citep{jannach2017user, bostandjiev2012tasteweights, swearingen2001beyond, schafer2002meta, schaffer2015hypothetical}.

@@

It is essential to base this study in a theoretical framework. As previously mentioned, we use Technology Acceptance Model by Davis \citeyearonly{davis1985technology}. Currently there are two dominant technology acceptance models used in the literature: Unified Theory of Acceptance and Use of Technology (UTAUT) and Technology Acceptance Model (TAM) \citep{venkatesh2003utaut, davis1985technology}. Both models provide valuable insights in understanding the users' acceptance of information systems. In the following section we will briefly discuss both TAM and UTAUT and argue why I have ultimately selected TAM for our study. Then, hypotheses will be developed.

@@

Because of its simplicity and versatility after introduction TAM has been used with a number of extensions and modifications to better suit specific research context and include relevant constructs \citep{amoako2004extension, burton2006mediationextension, featherman2003predictingextension}. Subsequently, those modifications have led to scholars building upon on TAM. TAM2 \citep{venkateshDeterminantsPerceivedEase2000} and TAM3 \citep{venkateshTechnologyAcceptanceModel2008}, UTAUT \citep{venkatesh2003utaut} have been introduced. Detailed review of TAM can be found in the study of Marangunic \citeyearonly{marangunic2015technology}.

@@

However, contrary to my expectations, H2 was not confirmed. Participants did not perceive the RS with control to be less easy to use. This result needs further discussions. The theoretical foundations for H2 have been driven from established research stream that established connection between information overload and decision making of consumers \citep{jacoby1974brand, chenEffectsInformationOverload2009}. While increased information leads to better informed decisions, research has established trade-off between information richness and decision quality \citep{malhotra1982information, jacoby1974brand}. Also, information overload has been found to negatively affect acceptance of internet utilization \citep{shih2004extended}.

@@

\subsection{Implications}

The growing speed of digitisation of our lives and growing AI presence means even more applications of recommender systems. As well made and diverse recommendations are crucial for the success of businesses and growing importance of human-AI interaction it is important to understand the challenges in implementation of control mechanisms \citep{songWhenHowDiversify2019, dietvorstEtAl18}. There are several important implications for system designers arising from this study.

@@

\subsection{Summary, limitations and future directions}

@@

The journey of my dissertation commenced with an enthusiastic aim of exploring the nature and nuanced influences of context effects on decision making and their applications on recommender system design. This voyage comprises four studies, each contributing to a richer understanding of these effects, and jointly offering a new perspective on the complex interplay of context effects.

@@

The approach taken in the second study mainly accommodated context effects as a result of ordinal relationships, overlooking the potential of cardinal measurements. Future research could tackle this by factoring in the magnitude of the relationship, adding depth to the analysis. Also, an interesting avenue could be the research towards quantifying context effects outside ``trinity".

Despite the effectiveness of the two-stage heuristics approach developed in the third study, its direct application to modern systems may not be straightforward due to traditional emphasis on predictive power. Albeit this emphasis has already started to gradually change, enhancing predictive capabilities this approach could definitely make it more attractive for integration into recommender systems.

Lastly, the fourth study has looked at the universe of user control mechanisms only through a small prism, employing two of them. However, there's a vast realm of other control mechanisms awaiting exploration in future research, calling for studies of more complex experimental settings, while maintaining domain neutrality. Albeit challenging, I consider this achievable.

In conclusion, the insights gleaned from this dissertation represent a lighthouse, illuminating the path towards a fascinating exploratory journey. The cumulative findings obtained from four studies provide a solid foundation for future research, paving the way for potential advancements in our understanding of context, its application on the design and operation of recommender systems, subsequently transforming how users engage with those tools throughout the digital landscape. As our exploration of context progresses, we will be ably to steadily unveil the substantial potential it possess. The research conducted and gathered in this dissertation signals promising prospects, fostering new insights and perspectives in the ever complex, dynamic relationship between decision sciences and digital technologies.

@@

1. Initialization: Generate an initial population of $NP$ candidate solutions randomly, each solution is a vector in the $D$-dimensional space.

@@

    $$X[i, j] = X_min[j] + rand(0, 1) * (X_max[j] - X_min[j])$$

@@

    $$V[i, j] = X[r1, j] + F * (X[r2, j] - X[r3, j])$$

@@

    $$
        U[i, j] =
        \begin{cases}
            V[i, j] & \text{if } \text{rand}(0,1) \leq \text{CR} \text{ or } j = j_{\text{rand}} \\
            X[i, j] & \text{otherwise}
        \end{cases}
    $$

@@

        $$
        X[i, j] =
        \begin{cases}
            U[i, j] & \text{if } f(U[i, j]) \leq f(X[i, j]) \\
            X[i, j] & \text{otherwise}
        \end{cases}
        $$

@@

$$\sum_{i=0}^{n}\min_{\mu_j\in C_j}\left(\left.\left|x_i-\mu_j\right|^2\right.\right)$$

>>>>

\title{{\bf \Large More than Meets the Eye: The Power of Context in Decision-Making and Recommendations}
}

@@

    This thesis demonstrates the existence of context effects in multi-attribute, multidimensional settings and develops methodologies of enhancing recommender systems design with context effects. Additionally, this study investigates external context, namely, the tools that allows users to express their preferences and how to make recommender systems better through them.

\end{abstract}
\newpage
\section{Introduction}

@@

Choice context has long received the attention of scholars in many fields, including marketing, psychology, economics. Because it was studied in many domains, scholars refer to it differently. This thesis concentrates and uses two of them. The first one posits that choice context \footnote{For the remainder of the dissertation, I will follow the previous literature \citep{truebloodEtAl13} and refer this definition of choice context as context effects.} is ``the availability and the nature of choice alternatives" \citep{tversky1972elimination, huberEtAl82, simonson89}. Previous research has demonstrated the existence of context effects in various settings \citep{herne1997decoy, soltani2012range, evangelidisEtAl18,  wuConsguner20}. The second definition of context arises much later with the proliferation of recommender systems. Context also referred to as ``the time and content of the choice, the location or socio-demographic characteristics of the decision-maker..." \citep{adomavicius2011context}\footnote{To distinguish between these two defintions I will refer to this definition as external context throughout this dissertation.}

@@

Research agrees on importance of metrics beyond accuracy for recommender systems \citep{kaminskas2016diversity}. It is in the best interest of online marketplaces to provide users with recommendations which are not only in the category ``exactly what I want", but also ``I never thought I would have liked this one" \citep{kotkovSurveySerendipityRecommender2016}. This can not only boost sales \citep{songWhenHowDiversify2019}, but also increase the satisfaction of users \citep{knijnenburgExplainingUserExperience2012}. Albeit, it was observed that the preferences of individuals are not stable and tend to change and system designers have proposed various tools to enable users signal their preference shifts \citep{bostandjiev2012tasteweights,hijikata2012relation}. However, to better understand it is necessary to study those tools in isolation from the context arising from the choice set and concentrate purely on external context that arises from the user side, e.g their preferences \citep{adomavicius2011context}. In my last study I address this gap by conducting an online experiment and applying Technology Acceptance Model \citep{davis1985technology} to measure users' acceptance of such tools.

@@

\section{Beyond the Experiment: Exploring Context Effects in Multi-Attribute, Multi-Alternative Choice Environments}\label{chapter:simulationStudy}
\begin{abstract}

    Previous computational decision making models developed to account for context effects have only been
    studied with an experimental data where only one effect was produced at a time. Using data coming from strictly controlled experimental environments
    hinders the understanding of context effects that occur in real-world choice scenarios where items have multiple dimensions and choice sets have dozens of alternatives. In this chapter
    I apply a computational model to an observational data which was not done before. The data comes from an air travel industry and is ideal to study context effects in multi-attribute, multia-lternative choice environments. I first find optimal parameters for computational model using differential evolution algorithm. Then, I complement a traditional choice model with its outputs and assess the significance of its contribution. This chapter contributes to context effect and decision making literature by providing further insights on behavior of computational decision making models in real-world choice data.

@@

\subsection{Big three context effects}\label{chapter:bigThreeContextEffectsDescription}

@@

\section{Unravelling Contextual Layers: Enhancing Choice Modeling in Multi-Attribute, Multi-Alternative Settings\footnote{This chapter is based on a joint work with my supervisor Zakaria Babutsidze, William Rand, Nobuyuki Hanaki, Ismael Rafai, Rodrigo Acuna Agost and Thierry Delahaye.}}\label{chapter:jmrPaper}

@@

Context effects have been systematically studied in marketing and psychology \citep{kivetz04, roodrkerkEtAl11, frederickEtAl14, dotsonEtAl18}. However, virtually all such studies have used controlled experiments in neat, compact settings. Namely, the settings where decision-makers are presented with few options and (very) few attributes across which these options differ. In contrast, most actual choices take place in much messier environments. Especially today when much of our search and shopping activity has shifted online. Proliferation of search engines allows each option to easily be compared with many alternatives across many different characteristics. In the chapter \ref{chapter:simulationStudy} I focused quantifying context as an aggregate. Although my previous study could successfully identify context in complex, multi-attribute setting,  the knowledge about the prevalence of context effects in these environments is still scarce. Precise measurements of context effects in multi-option and multi-attribute setting is one way to contribute. Defining how to measure these effects is a minimum requirement for proceeding to evaluate the existence of context effects using observational data.

There have been recent attempts in computer science to define some of these effects. Machine learning community has incorporated context effects in discrete choice models applied to observational data \citep{pfannschmidt2019learning, bowerBalzano20}. However, the aim therein is increased prediction accuracy of choice models \citep{tomlinsonBenson21}. As a result, incorporation of context effects takes the form of generalizing choice models to allow for departure from the strict rationality assumptions\footnote{Recent examples of this approach are  Contextual Multinomial Logit by Yusefi Maragheh et al. \citeyearonly{yousefi2020choice} and Linear Context Logit by Tomlinson and Benson \citeyearonly{tomlinsonBenson21}.}. These proposed generalizations of estimated functional forms usually do not distinguish across various different types of context effects. Additionally, these approaches often run into computational difficulties, i.e., the estimation process is NP-hard \citep{yousefi2020choice}.

In this chapter, I propose measures of different context effects in multi-option and multi-dimensional settings. Following Rooderkerk, Van Heerde, and Bijmolt \citeyearonly{roodrkerkEtAl11}, I consider three context effects - attraction, compromise and similarity effects. Although I have discussed each of these three effects in more detail in the chapter \ref{chapter:bigThreeContextEffectsDescription}, very brief recap of the "trinity" seems appropriate. Attraction effect refers to the increase in attractiveness of a set of option as a result of adding an alternative to the choice set, compromise effect refers to the inclination of consumers to prefer options that represent a compromise across extreme sets of alternatives, while similarity effect refers to the drop in choice likelihood for an alternative once another, similar, alternative has been added to the choice set. Each of the measures corresponding to three aforementioned effects requires a specific approach for making measurements applicable to observational data. Each of these measures is calculated prior to choice estimation, which avoids computational problems. After presenting the generalized measures of the three effects, we perform an empirical analysis of choices based on the new measures using observational data. We use an extensive dataset of airfare choices for this exercise. We identify that attraction and similarity effects do influence choices in air-travel booking data. We also detect a reverse compromise effect that seems to indicate that air-travelers consistently prefer extreme alternatives (i.e., the cheapest, or the shortest flight) to alternatives that constitute a compromise among extreme options.

@@

\newpage
\section{Tackling the Cold-Start Problem: A Context-Informed Approach to Recommender Systems\footnote{This chapter is based on a joint work with my supervisor Zakaria Babutsidze, William Rand and Thierry Delahaye. It has been published in the proceedings of 54th Hawaii International Conference on System Sciences}.}\label{chapter:hicssPaper}

@@

In this paper, I incorporate insights from the decision literature into recommender system design using the dataset of European flight choices \citep{lheritier2019airline}. I argue that the choice process occurs in two sequential stages: consumers first identify the small subset of the choices that they would ``consider" and then make a choice from that subset. Combining findings in marketing, management and consumer behavior, and using clustering to quantify the contextual information of the choice set,  I propose a user-side, two-step  ``consider-then-choose" \citep{gilbride2004choice,liu2011efficient} approach to recommender system design to tackle the cold-start problem.

@@

Elimination by aspects is another heuristic approach which has been
proposed in the literature \citep{tversky1972elimination}. The basic setup of this approach is that an individual chooses one attribute, and eliminates options based on this attribute and repeats this procedure for other attributes if necessary until the remaining options do not share common attributes anymore. Then, as a last step, the final option is chosen according to Luce’s choice axiom \citep{luce2012individual} which states that the probability of selecting one option over the others in a choice set is not affected by presence or absence of other options. Most of the results on this topic \citep{batsell1985new,gensch1987two,currim1988disaggregate,manrai1989elimination} indicate that the use of multi-phase heuristic processes can increase the accuracy of the estimation and result in improved interpretability of the models. Elimination by aspects is considered a heuristic method with stochastic rules because of the nature of the comparisons an individual makes, and because the selection process is not based on the relative importance of the features 
\citep{aribarg2018advancing}. Such models are also hard to apply successfully because they require tremendous numbers of parameters to be estimated \citep{batsell1985new}. Although these models can theoretically capture the essence of the two-stage choice process, they are not able to identify the results of separate stages \citep{gilbride2004choice}.

@@

For an empirical study of consideration set formation, one can elicit information on consideration sets in multiple ways \citep{gaskin2007two,yee2007greedoid, ding2011unstructured}. However, for modeling purposes the literature discusses two main ways of consideration set formation: deterministic \citep{coombs1951mathematical} and stochastic \citep{mcfadden1973conditional, urban1984testing}. While stochastic modeling makes all potential sets possible by attaching non-zero choice probability to each of them, deterministic approaches may render some outcomes impossible \citep{aribarg2018advancing}. Because we can not know which answer is the best and the decision-maker, or consumer, is the final arbiter of the ``correct" choice \citep{hauser2014consideration}, the use of either of these two approaches must consider the choice environment, time frame, future value (or loss) associated with the correct and incorrect choice, and so on \citep{punj2009information}. For example, let's consider the flight booking case again and suppose that the consumer lives far from the airport and can reach it only in the afternoon. Consequently, all tickets with departure time before midday would not be considered at all. When forming a consideration set in this case, one must not only consider the characteristics of available options, but also the characteristics of the consumer and the choice environment. While using a purely stochastic approach might yield sets which include some options that consumer would indeed consider, there will also be options which will have zero probability for consideration. In contrast, applying some deterministic rules derived from this particular choice environment, such as the departure time, in the consideration set formation, will exclude those options completely.

When there are not many options to consider, options have few attributes, or final choice utility is not evenly distributed among the attributes, the consideration sets may be modelled via simple deterministic rules, because there is not much cognitive load and the decision rules are relatively simple  \citep{lee2004effect, hauser2014consideration} . When forming consideration sets, it is also important to consider their size. It is very difficult to decide an optimal size based on the choice environment and individual processing capabilities \citep{de2011modelling}.

@@

@@

In this study, I demonstrated the importance of context effects derived from choice sets in shaping user decisions, as well as the feasibility of incorporating these effects into the design of recommender systems.
Albeit enhancing the system's ability to align its recommendations on the account of user's two-step decision making process is a significant step forward, one must also recognize that context does not arise from the choice set alone. In previous chapters I have used Tversky's \citeyearonly{tversky1972elimination} definition of context being ``the composition  and the nature of the choice set, availability of various options in it". More recent stream of research has also established that choice context is also dependent on user preferences among other things \citep{dey2001understanding, adomavicius2011context}.

@@

\section{Personalization in Context: User Control and Its Impact on Recommender System Acceptance}\label{chapter:UserControlAndRS}
\begin{abstract}

@@

The importance of precisely defining the concepts and constructs used in this study is crucial. Defining the key constructs not only establishes a common understanding, but also provides boundary conditions and ensures the reliability of the research findings \citep{creswell2013research}.

By clearly defining RS one is able to distinguish among related concepts such as recommendation agents (RA) and interactive decision aid tools (IDA) which wile sharing some commonalities, also entail some important differences in their goals, functionalities and user interactions \citep{xiao2007commerce}. By specifying the characteristics of RS in our particular context, I aim to ensure the relevance of the findings. This is also important considering the methodology and framework.

@@

Henceforth, in the following subsections, I will discuss the definitions of recommender systems and user control in detail and justify our choices for this study.

@@

The second stage of user control entails functionalities which allow users to influence recommendations after they are generated. The previous approaches have user sorting, feature and weight based filtering and critiquing. \citep{swearingen2001beyond, schafer2002meta, bostandjiev2012tasteweights, schaffer2015hypothetical, jannach2017user}.

@@

It is essential to base this study in a theoretical framework. As previously mentioned, we use Technology Acceptance Model by Davis \citeyearonly{davis1985technology}. Currently there are two dominant technology acceptance models used in the literature: Unified Theory of Acceptance and Use of Technology (UTAUT) and Technology Acceptance Model (TAM) \citep{davis1985technology, venkatesh2003utaut}. Both models provide valuable insights in understanding the users' acceptance of information systems. In the following section we will briefly discuss both TAM and UTAUT and argue why I have ultimately selected TAM for our study. Then, hypotheses will be developed.

@@

Because of its simplicity and versatility after introduction TAM has been used with a number of extensions and modifications to better suit specific research context and include relevant constructs \citep{featherman2003predictingextension, amoako2004extension, burton2006mediationextension}. Subsequently, those modifications have led to scholars building upon on TAM. TAM2 \citep{venkateshDeterminantsPerceivedEase2000} and TAM3 \citep{venkateshTechnologyAcceptanceModel2008}, UTAUT \citep{venkatesh2003utaut} have been introduced. Detailed review of TAM can be found in the study of Marangunic \citeyearonly{marangunic2015technology}.

@@

However, contrary to my expectations, H2 was not confirmed. Participants did not perceive the RS with control to be less easy to use. This result needs further discussions. The theoretical foundations for H2 have been driven from established research stream that established connection between information overload and decision making of consumers \citep{jacoby1974brand, chenEffectsInformationOverload2009}. While increased information leads to better informed decisions, research has established trade-off between information richness and decision quality \citep{jacoby1974brand, malhotra1982information}. Also, information overload has been found to negatively affect acceptance of internet utilization \citep{shih2004extended}.

@@

\textbf{Implications}

The growing speed of digitisation of our lives and growing AI presence means even more applications of recommender systems. As well made and diverse recommendations are crucial for the success of businesses and growing importance of human-AI interaction it is important to understand the challenges in implementation of control mechanisms \citep{dietvorstEtAl18, songWhenHowDiversify2019}. There are several important implications for system designers arising from this study.

@@

\textbf{Summary, limitations and future directions}

@@

The journey of my dissertation commenced with an enthusiastic aim of exploring the nature and nuanced influences of context effects on decision making and their applications on recommender system design. This voyage comprises of four studies, each contributing to a richer understanding of these effects, and jointly offering a new perspective on the complex interplay of context effects.

@@

The approach taken in the second study mainly accommodated context effects as a result of ordinal relationships, overlooking the potential of cardinal measurements. Future research could tackle this by factoring in the magnitude of the relationship, adding depth to the analysis. Also, an interesting avenue could be the research towards quantifying context effects outside the ``trinity".

Despite the effectiveness of the two-stage heuristics approach developed in the third study, its direct application to modern systems may not be straightforward due to traditional emphasis on predictive power. Albeit this emphasis has already started to gradually change, enhancing predictive capabilities of this approach could definitely make it more attractive for integration into recommender systems.

Lastly, the fourth study has looked at the universe of user control mechanisms only through a small prism, employing two of them. However, there's a vast realm of other control mechanisms awaiting exploration in future research, calling for studies of more complex experimental settings, while maintaining domain neutrality. Notwithstanding the fact that it seems challenging, I consider this achievable.

In conclusion, the insights gleaned from this dissertation represent a lighthouse, illuminating the path towards a fascinating exploratory journey that lies ashore. The cumulative findings obtained from four studies provide a solid foundation for future research, paving the way for potential advancements in our understanding of context, its application on the design and operation of recommender systems, subsequently transforming how users engage with those tools throughout the digital landscape. As our exploration of context progresses, we will be ably to steadily unveil the substantial potential it possess. The research conducted and gathered in this dissertation signals promising prospects, fostering new insights and perspectives in the ever complex, dynamic relationship between decision sciences and digital technologies.

@@

1. Initialization: Generate an initial population of $NP$ candidate solutions randomly, each solution is a vector in the $D$-dimensional space

@@

    $$X[i, j] = X_min[j] + rand(0, 1) * (X_max[j] - X_min[j]) .$$

@@

    $$V[i, j] = X[r1, j] + F * (X[r2, j] - X[r3, j]) .$$

@@

    $$
        U[i, j] =
        \begin{cases}
            V[i, j] & \text{if } \text{rand}(0,1) \leq \text{CR} \text{ or } j = j_{\text{rand}} \\
            X[i, j] & \text{otherwise}
        \end{cases} .
    $$

@@

        $$
        X[i, j] =
        \begin{cases}
            U[i, j] & \text{if } f(U[i, j]) \leq f(X[i, j]) \\
            X[i, j] & \text{otherwise}
        \end{cases} .
        $$

@@

$$\sum_{i=0}^{n}\min_{\mu_j\in C_j}\left(\left.\left|x_i-\mu_j\right|^2\right.\right) .$$
